---
title: "Effect of Truth-in-Sentencing on Incarceration Rate"
author: "Michael Ippolito"
date: "2023-03-26"
output:
  pdf_document: default
  html_document: default
subtitle: CUNY Data608 Final Project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(gridExtra)
```

## Introduction

Before the 1990s, many prisoners incarcerated in the US criminal justice system only served a fraction of their sentence, often being released for good behavior or other mitigating factors. These factors were applied unevenly at the discretion of parole boards. Public pressure led to the passage of the Violent Crime Control and Law Enforcement Act (VCCLEA) in 1994, which gave incentives to states that required prisoners to serve at least 85% of their sentence.

Arguments leading to passage of the VCCEA included that truth in sentencing (TIS) laws would deter future criminal behavior, would render justice to the perpetrator for the crime commited, and that time served would theoretically be unbiased and no longer at the whim of a parole board. Opponents of TIS argue that TIS leads to prison overcrowding, that the cost of incarcerating prisoners longer could be better spent on crime prevention programs, that it disincentivizes good behavior, and that it in fact does not serve as a deterrent to future crime. Further, it leaves judges no room in weighing extenuating factors when rendering a sentence.

My project will seek to evaluate whether TIS laws impact the incarceration rate, and whether other factors may be at play. The data source I'll be using comes from the Inter-university Consortium for Political and Social Research (ICPSR), a consortium of almost 800 institutions contributing and sharing data on a wide array of social and behavioral science topics. The data is linked here:

https://www.icpsr.umich.edu/web/NACJD/studies/4456

While the data was collected from 1972 to 2002, it is still relevant as TIS laws are still on the books in the U.S. Using an interactive Dash app deployed via Heroku, I plan on showing how the incarceration rate changes based on the extent to which each U.S. state enforces TIS, whether mandatory "enhancements" to sentences affect the rate, and whether any non-judiciary factors (e.g. poverty rate) might also have an effect on the rate.


## Data Exploration

```{r}

# Read data
dfvar = read.csv('data/vars.tsv', sep='\t')
dfabbr = read.csv('data/state_abbr.csv')
df1 = read.csv('data/04456-0001-Data.tsv', sep='\t')
df2 = read.csv('data/04456-0002-Data.tsv', sep='\t')

```

```{r}

# Summary - outcome vars
print("")
print("Summary - outcome vars")
summary(df1 %>%
    select(
        inc_rate,
        lag_incr,
        d_inc_ra,
        admissio,
        adm_100k,
        ncourtc,
        court_10,
        pro_drug,
        drug_arr,
        cri_l1,
        vio_l1,
        pro_crim
    )
)

# Missing values

# inc_rate: none missing
print(paste('inc_rate', 'missing',0))

# lag_incr: missing=999
print(paste('lag_incr', 'missing', nrow(df1[df1['lag_incr'] == 999,])))

# d_inc_ra: missing=999
print(paste('d_inc_ra', 'missing', nrow(df1[df1['d_inc_ra'] == 999,])))

# admissio: missing=999999
print(paste('admissio', 'missing', nrow(df1[df1['admissio'] == 999999,])))

# adm_100k: missing=999
print(paste('adm_100k', 'missing', nrow(df1[df1['adm_100k'] == 999,])))

# ncourtc: missing=99999
print(paste('ncourtc', 'missing', nrow(df1[df1['ncourtc'] == 99999,])))

# court_10: missing=999
print(paste('court_10', 'missing', nrow(df1[df1['court_10'] == 999,])))

# drug_arr: missing=9999
print(paste('drug_arr', 'missing', nrow(df1[df1['drug_arr'] == 9999,])))

# pro_drug: missing=9
print(paste('pro_drug', 'missing', nrow(df1[df1['pro_drug'] == 9,])))

# cri_l1: none missing
print(paste('cri_l1', 'missing',0))

# vio_l1: none missing
print(paste('vio_l1', 'missing', 0))

# pro_crim: none missing
print(paste('pro_crim', 'missing',0))

```

```{r}

# Histograms - incarceration rate
fig1 <- plot_ly(type='histogram', x=df1$inc_rate, nbinsx=40) %>%
    layout(title='Incarceration Rate', xaxis=list(title='Incarceration rate per 100,000 residents'))
fig2 <- plot_ly(type='histogram', x=df1[df1$lag_incr != 999, 'lag_incr'], nbinsx=40) %>%
    layout(title='Lagged Incarceration Rate', xaxis=list(title='Incarceration rate per 100,000 residents (lagged one year)'))
fig3 <- plot_ly(type='histogram', x=df1[df1$d_inc_ra != 999, 'd_inc_ra'], nbinsx=40) %>%
    layout(title='Incarceration Rate Difference', xaxis=list(title='Difference in incarceration rate (t1 - t0)'))
grid.arrange(fig1, fig2, fig3)

```

```{r}

# Histograms - prison admissions
fig <- plot_ly(type='histogram', x=df1[df1$admissio != 999999, 'admissio'], nbinsx=40) %>%
    layout(title='Prison Admissions', xaxis=list(title='Number of prison admissionss'))
fig
fig <- plot_ly(type='histogram', x=df1[df1$adm_100k != 999, 'adm_100k'], nbinsx=40) %>%
    layout(title='Prison Admission Rate', xaxis=list(title='Prison admissions per 100,000 residents'))
fig

```

```{r}

# Histograms - court commitments
fig <- plot_ly(type='histogram', x=df1[df1$ncourtc != 99999, 'ncourtc'], nbinsx=40) %>%
    layout(title='Court Commitments', xaxis=list(title='Nubmer of court commitments'))
fig
fig <- plot_ly(type='histogram', x=df1[df1$court_10 != 999, 'court_10'], nbinsx=40) %>%
    layout(title='Court Commitment Rate', xaxis=list(title='Court commitment rate per 100,000 residents'))
fig

```

```{r}

# Histograms - drug arrests
fig <- plot_ly(type='histogram', x=df1[df1$pro_drug != 9, 'pro_drug'], nbinsx=40) %>%
    layout(title='Drug Arrests', xaxis=list(title='Three-year average rate of drug arrests (drug arrests/total arrests)'))
fig
fig <- plot_ly(type='histogram', x=df1[df1$drug_arr != 9999, 'drug_arr'], nbinsx=40) %>%
    layout(title='Drug Arrest Rate', xaxis=list(title='Drug arrest rate per capita'))
fig

```

```{r}

# Histograms - crime rates
fig <- plot_ly(type='histogram', x=df1$cri_l1, nbinsx=40) %>%
    layout(title='Crime Rate', xaxis=list(title='Crime rate per 100,000 residents (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$vio_l1, nbinsx=40) %>%
    layout(title='Violent Crime Rate', xaxis=list(title='Violent crime rate per 100,000 residents (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$pro_crim, nbinsx=40) %>%
    layout(title='Property crime Rate', xaxis=list(title='Property crime rate per 100,000 residents (lagged one year)'))
fig


```

```{r}

# Factor
df1$state_na = factor(df1$state_na)
df1$south = factor(df1$south)
df1$west = factor(df1$west)
df1$east = factor(df1$east)
df1$midwest = factor(df1$midwest)
df1$year = factor(df1$year)
df1$govern = factor(df1$govern)

# Summary - control vars
print("")
print("Summary - control vars")
summary(df1 %>%
    select(
        state_na,
        south,
        west,
        east,
        midwest,
        year,
        pop_l1,
        incpc_l1,
        unemp_l1,
        p_1824_l,
        p_25_34_,
        p_black_,
        p_hisp_l,
        metro_l1,
        poverty_,
        expcorr_,
        correxp_,
        rev_l2_1,
        wel_l1_1,
        religion,
        gini,
        pol_100k,
        ideo2,
        gov2,
        govern
    )
)

# Missing values
# Skipped states and years
# No missing values for any var except:

# p_black_: missing=99
print(paste('p_black_', 'missing', nrow(df1[df1['p_black_'] == 99,])))

```

```{r}

# Histograms - demographics
fig <- plot_ly(type='histogram', x=df1$pop_l1, nbinsx=40) %>%
    layout(title='Population', xaxis=list(title='Population (thousands - lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$p_1824_l, nbinsx=40) %>%
    layout(title='Population 18-24', xaxis=list(title='Pct of population 18-24 (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$p_25_34_, nbinsx=40) %>%
    layout(title='Population 25-34', xaxis=list(title='Pct of population 25-34 (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1[df1$p_black_ != 99, 'p_black_'], nbinsx=40) %>%
    layout(title='Black Population', xaxis=list(title='Pct of population Black (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$p_hisp_l, nbinsx=40) %>%
    layout(title='Hispanic Population', xaxis=list(title='Pct of population Hispanic (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$metro_l1, nbinsx=40) %>%
    layout(title='Metropolitan Population', xaxis=list(title='Pct of population living in metropolitan areas (lagged one year)'))
fig

```

```{r}

# Histograms - economy
fig <- plot_ly(type='histogram', x=df1$incpc_l1, nbinsx=40) %>%
    layout(title='Income', xaxis=list(title='Income per capita (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$poverty_, nbinsx=40) %>%
    layout(title='Poverty Rate', xaxis=list(title='Poverty rate (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$unemp_l1, nbinsx=40) %>%
    layout(title='Unemployment Rate', xaxis=list(title='Unemployment rate (lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$rev_l2_1, nbinsx=40) %>%
    layout(title='State Revenues', xaxis=list(title='State revenues per 100,000 residents (constant US$ - lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$wel_l1_1, nbinsx=40) %>%
    layout(title='Welfare Expenditures', xaxis=list(title='Welfare expenditures per 100,000 residents (constant US$ - lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$gini, nbinsx=40) %>%
    layout(title='Gini Coefficient', xaxis=list(title='Gini coefficient for families (lower=more equally distributed income)'))
fig

```

```{r}

# Histograms - corrections
fig <- plot_ly(type='histogram', x=df1$expcorr_, nbinsx=40) %>%
    layout(title='Corrections Expenditures', xaxis=list(title='Direct expenditures on corrections (constant US$ - lagged one year)'))
fig
fig <- plot_ly(type='histogram', x=df1$correxp_, nbinsx=40) %>%
    layout(title='Corrections Expenditures Per Capita', xaxis=list(title='Corrections Expendutres per capita'))
fig
fig <- plot_ly(type='histogram', x=df1$pol_100k, nbinsx=40) %>%
    layout(title='Police Officers', xaxis=list(title='Full-time equivaluent police officers per 100,000 residents'))
fig

```

```{r}

# Histograms - society
fig <- plot_ly(type='histogram', x=df1$religion, nbinsx=40) %>%
    layout(title='Fundamentalist Religion', xaxis=list(title='Pct of population having fundamentalist religion'))
fig
fig <- plot_ly(type='histogram', x=df1$ideo2, nbinsx=40) %>%
    layout(title='Citizen Ideology', xaxis=list(title='Citizen ideology scale (lagged one year - higher=more liberal)'))
fig
fig <- plot_ly(type='histogram', x=df1$gov2, nbinsx=40) %>%
    layout(title='Government Ideology', xaxis=list(title='Government ideology scale (lagged one year - higher=more liberal)'))
fig
fig <- plot_ly(type='histogram', x=df1$govern, nbinsx=40) %>%
    layout(title='Republican Governors', xaxis=list(title='Republican Governors (1=yes)'))
fig

```

```{r}

# Factor
df1$tiss = factor(df1$tiss)
df1$tisviol = factor(df1$tisviol)
df1$pre_guid <- factor(df1$pre_guid)
df1$pre_volg <- factor(df1$pre_volg)
df1$pre_par <- factor(df1$pre_par)
df1$rec_sent <- factor(df1$rec_sent)
df1$presum_s <- factor(df1$presum_s)
df1$pre_rec_ <- factor(df1$pre_rec_)

# Summary - sentencing vars
print("")
print("Summary - sentencing vars")
summary(df1 %>%
    select(
        tisp,
        tiss,
        tisviol,
        pre_guid,
        pre_volg,
        pre_par,
        rec_sent,
        presum_s,
        pre_rec_
    )
)

# tisp: missing=999
print(paste('tisp', 'missing', nrow(df1[df1['tisp'] == 999,])))

# tiss: missing=9
print(paste('tiss', 'missing', nrow(df1[df1['tiss'] == 9,])))

# tisviol: missing=9
print(paste('tisviol', 'missing', nrow(df1[df1['tisviol'] == 9,])))

```

```{r}

# Histograms - tiss policy
fig <- plot_ly(type='histogram', x=df1[df1$tisp != 999, 'tisp'], nbinsx=40) %>%
    layout(title='TIS Policy - % of Sentence', xaxis=list(title='Pct of sentence imposed most offenders required to serve'))
fig
fig <- plot_ly(type='histogram', x=df1[df1$tiss != 9, 'tiss'], nbinsx=40) %>%
    layout(title='TIS Policy - Min Time Served Requirement', xaxis=list(title='Minimum time served requirement present'))
fig
fig <- plot_ly(type='histogram', x=df1[df1$tisviol != 9, 'tisviol'], nbinsx=40) %>%
    layout(title='TIS Policy - Violent Offenses', xaxis=list(title='Separate time served requred for violent offenses'))
fig

```

```{r}

fig <- plot_ly(type='histogram', x=df1$pre_guid, nbinsx=40) %>%
    layout(title='Presumptive Guidelines', xaxis=list(title='Presence of presumptive guidelines'))
fig
fig <- plot_ly(type='histogram', x=df1$pre_volg, nbinsx=40) %>%
    layout(title='Voluntary Guidelines', xaxis=list(title='Presence of voluntary guidelines'))
fig
fig <- plot_ly(type='histogram', x=df1$pre_par, nbinsx=40) %>%
    layout(title='Determinate Sentencing', xaxis=list(title='Presence of determinate sentencing'))
fig
fig <- plot_ly(type='histogram', x=df1$rec_sent, nbinsx=40) %>%
    layout(title='Recommended Sentences', xaxis=list(title='Presence of recommended sentences'))
fig
fig <- plot_ly(type='histogram', x=df1$presum_s, nbinsx=40) %>%
    layout(title='Presumptive Sentences', xaxis=list(title='Presence of presumptive sentences'))
fig
fig <- plot_ly(type='histogram', x=df1$pre_rec_, nbinsx=40) %>%
    layout(title='Presumptive Recommneded Sentences', xaxis=list(title='Presence of presumptive recommended sentences'))
fig

```

```{r}

# Summary - enhancement vars
print("")
print("Summary - enhancement vars")
summary(df1 %>%
    select(
        m1_weap,
        m2_prox,
        m3_offic,
        m4_harm,
        m5_cont,
        m6_citi,
        m7_repe,
        m8_mult,
        m9_drug
    )
)

```

```{r}

# Summary - mandatory sentences with weapon as trigger
print("")
print("Summary - mandatory sentences with weapon as trigger")
summary(df1 %>%
    select(
        m10_we1a,
        m11_we1b,
        m12_we1c,
        m13_we1d,
        m14_we1e,
        m15_we1f,
        m16_we2a,
        m17_we2b,
        m18_we2c,
        m19_we2d,
        m20_we2e,
        m21_we2f,
        m22_we3a,
        m23_we3b,
        m24_we3c,
        m25_we3d,
        m26_we3e,
        m27_we3f,
        m28_we4a,
        m29_we4b,
        m30_we4c,
        m31_we4d,
        m32_we4e,
        m33_we4f
    )
)

```

```{r}

# Summary - mandatory sentences with drugs as trigger
print("")
print("Summary - mandatory sentences with drugs as trigger")
summary(df1 %>%
    select(
        m34_dr1a,
        m35_dr1b,
        m36_dr1c,
        m37_dr1d,
        m38_dr1e,
        m39_dr1f,
        m40_dr2a,
        m41_dr2b,
        m42_dr2c,
        m43_dr2d,
        m44_dr2e,
        m45_dr2f,
        m46_dr3a,
        m47_dr3b,
        m48_dr3c,
        m49_dr3d,
        m50_dr3e,
        m51_dr3f
    )
)

```

```{r}

# Summary - mandatory sentences with bodily harm as trigger
print("")
print("Summary - mandatory sentences with bodily harm as trigger")
summary(df1 %>%
    select(
        m52_vi1a,
        m53_vi1b,
        m54_vi1c,
        m55_vi1d,
        m56_vi1e,
        m57_vi1f,
        m58_vi2a,
        m59_vi2b,
        m60_vi2c,
        m61_vi2d,
        m62_vi2e,
        m63_vi2f
    )
)

```


```{r}

df1 %>%
    filter(pre_rec_ == 1) %>%
    group_by(year) %>%
    summarize(n=n()) %>%
    ungroup() %>%
    arrange(year)

df1 %>%
    filter(tiss == 1) %>%
    group_by(year) %>%
    summarize(n=n()) %>%
    ggplot() +
    geom_point(aes(x=year, y=n)) +
    theme_minimal() +
    ggtitle('Number of states with TIS policy')

df1 %>%
    filter(tisp != 999) %>%
    group_by(year) %>%
    summarize(n=mean(tisp)) %>%
    ggplot() +
    geom_point(aes(x=year, y=n)) +
    theme_minimal() +
    ggtitle('Mean time served requirement (pct of sentnece)')


df1 %>%
    filter(tiss == 1) %>%
    group_by(year) %>%
    summarize(n=n()) %>%
    ggplot() +
    geom_point(aes(x=year, y=n)) +
    theme_minimal() +
    ggtitle('Number of states with TIS policy')

```

Sentencing	TIS policy	tisp	Pct sentence imposed most offenders required to serve
Sentencing	TIS policy	tiss	Min time served req'd
Sentencing	TIS policy	tisviol	Separate time served req for violent offenses
Sentencing	TIS policy	tisviol	Separate time served req for violent offenses
Sentencing	Guidelines	pre_guid	Presence of presumptive guidelines (guidelines contained in legislation)
Sentencing	Guidelines	pre_volg	Presence of voluntary guidelines (judges only use sentences as a guideline but can depart from them)
Sentencing	Guidelines	pre_par	Presence of determinate sentencing (sentences have a defined length)
Sentencing	Guidelines	rec_sent	Presence of recommended sentences (assume this means indeterminate sentencing, i.e. has a range)
Sentencing	Guidelines	presum_s	Presence of presumptive sentences (sentences contained in legislation)
Sentencing	Guidelines	pre_rec_	Presence of presumptive recommended sentences (assume this means ranges are contained in legislation)

```{r}

vars <- c('tiss', 'tisviol', 'pre_guid', 'pre_volg', 'pre_par', 'rec_sent', 'presum_s', 'pre_rec_')
dfyr <- df1 %>%
    select(year) %>%
    group_by(year) %>%
    summarize() %>%
    ungroup()
df3 <- dfyr

for (v in vars) {

    dftmp <- df1[df1[v] == 1,]
    dftmp <- dftmp %>%
        #filter(v == 1) %>%
        group_by(year) %>%
        summarize(n=n()) %>%
        ungroup()
    dftmp[v] = dftmp['n']
    dftmp <- dftmp %>% select(-n)
    df3 <- df3 %>%
        merge(dftmp, by='year', all.x=T)
}
df3 <- gather(df3, key='stat', val='val', -year)

df3 %>%
    filter(year !=1972) %>%
    ggplot() +
    geom_bar(aes(x=year, y=val, fill=stat), stat='identity', position = 'dodge') +
    theme_minimal()

df4 <- df1 %>%
    group_by(year, tiss) %>%
    summarize(n=n(), mean_tisp=mean(tisp))

df5 <- df1 %>%
    filter(year == 1999) %>%
    select(state_na, tiss, tisviol, tisp, pre_guid, pre_volg, pre_par, rec_sent, presum_s, pre_rec_)

df6 <- df1 %>%
    mutate(fed=ifelse(tisp>=85, 1, 0)) %>%
    group_by(year, fed) %>%
    summarize(n=n()) %>%
    ungroup() %>%
    spread(key=fed, value=n)

df7 <- df1 %>%
    mutate(tmpviol=ifelse(m10_we1a==1 | m16_we2a==1 | m22_we3a==1 | m28_we4a==1 | m52_vi1a==1 | m58_vi2a==1, 1, 0)) %>%
    group_by(year, tmpviol) %>%
    summarize(n=n()) %>%
    ungroup() %>%
    spread(key=tmpviol, value=n, sep='=')

df7b <- df1 %>%
    #filter(tisp != 0) %>%
    mutate(tmpviol=ifelse(m10_we1a==1 | m16_we2a==1 | m22_we3a==1 | m28_we4a==1 | m52_vi1a==1 | m58_vi2a==1, 1, 0)) %>%
    #mutate(tmpmo=mean(m15_we1f) + mean(m21_we2f) + mean(m27_we3f) + mean(m33_we4f) + mean(m57_vi1f) + mean(m63_vi2f)) %>%
    mutate(tmpagg=ifelse(pre_guid==1 | pre_volg==1 | pre_par==1 | rec_sent==1 | presum_s==1 | pre_rec_==1, 1, 0)) %>%
    group_by(year, pre_par) %>%
    summarize(meantisp=mean(tisp), .groups='keep') %>%
    ungroup() %>%
    spread(key=pre_par, value=meantisp, sep='=')

```

## References

Stemen, D. _Impact of State Sentencing Policies on Incarceration Rates in the United States, 1975-2002 (ICPSR 4456)._ (2007, September 27). Inter-university Consortium for Political and Social Research. https://www.icpsr.umich.edu/web/NACJD/studies/4456

Ditton, P., Wilson, D. & BJS Statisticians. _Truth in Sentencing in State Prisons._ (1999, January). Bureau of Justice. https://bjs.ojp.gov/content/pub/pdf/tssp.pdf

Shorey, J. _Truth in Sentencing Overview & Laws._ (2022, May 4). Study.com. https://study.com/learn/lesson/truth-in-sentencing-overview-laws.html

_State Good Time and Earned Time Laws (2021, June 11)._ National Conference of State Legislatures. https://www.ncsl.org/civil-and-criminal-justice/state-good-time-and-earned-time-laws

_Know More: Truth-in-Sentencing._ (2020). Restore Justice Foundation. https://www.restorejustice.org/about-us/resources/know-more/know-more-truth-in-sentencing/

_Federal Sentencing Guidelines: Background, Legal Analysis, and Policy Options_ (2009, March 16). EveryCRSReport. https://www.everycrsreport.com/reports/RL32766.html

_Metadata Glossary_ (2023). World Bank, https://databank.worldbank.org/metadataglossary/gender-statistics/series/SI.POV.GINI.
