---
title: "Effect of Truth-in-Sentencing on Incarceration Rate"
author: "Michael Ippolito"
date: "2023-03-26"
output:
  pdf_document: default
  html_document: default
subtitle: CUNY Data608 Final Project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
```

## Final Project Writeup

In my project, I explored the effects of truth-in-sentencing (TIS) laws on incarceration rates and rates of violent crime. I chose to use a data source from the National Archive of Criminal Justice Data at the University of Michigan, collected from 1972 to 2002 as part of a study specifically designed to evaluate the effects of truth-in-sentencing.

The data included 298 variables and 550 observations (11 years of data for each of the 50 US states) with a very low proportion of missing values. I limited my outcome variables to incarceration rate and violent crime rate. The remaining variables included a wide range of factors that may influence incarceration and crime rates, such as the presence of TIS laws, whether determinate or indeterminate sentencing was in use, so-called "enhancements" (extra prison time) for certain offenses, and the percentage of their sentences offenders are required to serve. The data also includes control variables such as economic factors (poverty rate and Gini coefficient), social factors (whether the state has a Republican governor and what percentage of the population observe a fundamentalist religion), and how much is spent per capita on corrections and law enforcement.

What I observed from the data is that TIS can be interpreted to be either working or not working, depending on your ideology. If the goal is to reduce violent crime rates, it does seem to be having that effect:

```{r echo=FALSE, include=FALSE}

# Read data
dfvar = read.csv('data/vars.tsv', sep='\t')
dfabbr = read.csv('data/state_abbr.csv')
df1 = read.csv('data/04456-0001-Data.tsv', sep='\t')
df2 = read.csv('data/04456-0002-Data.tsv', sep='\t')

```

```{r echo=FALSE, include=FALSE}

# Factor
#df1$state_na = factor(df1$state_na)
df1$south = factor(df1$south)
df1$west = factor(df1$west)
df1$east = factor(df1$east)
df1$midwest = factor(df1$midwest)
#df1$year = factor(df1$year)
df1$govern = factor(df1$govern)
df1$tiss = factor(df1$tiss)
df1$tisviol = factor(df1$tisviol)
df1$pre_guid <- factor(df1$pre_guid)
df1$pre_volg <- factor(df1$pre_volg)
df1$pre_par <- factor(df1$pre_par)
df1$rec_sent <- factor(df1$rec_sent)
df1$presum_s <- factor(df1$presum_s)
df1$pre_rec_ <- factor(df1$pre_rec_)

```

```{r echo=FALSE, include=FALSE}

vars <- c('tiss', 'tisviol', 'pre_guid', 'pre_volg', 'pre_par', 'rec_sent', 'presum_s', 'pre_rec_')
dfyr <- df1 %>%
    dplyr::select(year) %>%
    group_by(year) %>%
    summarize() %>%
    ungroup()
df3 <- dfyr

for (v in vars) {

    dftmp <- df1[df1[v] == 1,]
    dftmp <- dftmp %>%
        #filter(v == 1) %>%
        group_by(year) %>%
        summarize(n=n()) %>%
        ungroup()
    dftmp[v] = dftmp['n']
    dftmp <- dftmp %>% dplyr::select(-n)
    df3 <- df3 %>%
       merge(dftmp, by='year', all.x=T)
}
df3 <- gather(df3, key='stat', val='val', -year)

df3 %>%
    filter(year !=1972) %>%
    ggplot() +
    geom_bar(aes(x=year, y=val, fill=stat), stat='identity', position = 'dodge') +
    theme_minimal()

df4 <- df1 %>%
    group_by(year, tiss) %>%
    summarize(n=n(), mean_tisp=mean(tisp))

df5 <- df1 %>%
    filter(year == 1999) %>%
    dplyr::select(state_na, tiss, tisviol, tisp, pre_guid, pre_volg, pre_par, rec_sent, presum_s, pre_rec_)

df6 <- df1 %>%
    mutate(fed=ifelse(tisp>=85, 1, 0)) %>%
    group_by(year, fed) %>%
    summarize(n=n()) %>%
    ungroup() %>%
    spread(key=fed, value=n)

df7 <- df1 %>%
    mutate(tmpviol=ifelse(m10_we1a==1 | m16_we2a==1 | m22_we3a==1 | m28_we4a==1 | m52_vi1a==1 | m58_vi2a==1, 1, 0)) %>%
    group_by(year, tmpviol) %>%
    summarize(n=n()) %>%
    ungroup() %>%
    spread(key=tmpviol, value=n, sep='=')

df7b <- df1 %>%
    #filter(tisp != 0) %>%
    mutate(tmpviol=ifelse(m10_we1a==1 | m16_we2a==1 | m22_we3a==1 | m28_we4a==1 | m52_vi1a==1 | m58_vi2a==1, 1, 0)) %>%
    #mutate(tmpmo=mean(m15_we1f) + mean(m21_we2f) + mean(m27_we3f) + mean(m33_we4f) + mean(m57_vi1f) + mean(m63_vi2f)) %>%
    mutate(tmpagg=ifelse(pre_guid==1 | pre_volg==1 | pre_par==1 | rec_sent==1 | presum_s==1 | pre_rec_==1, 1, 0)) %>%
    group_by(year, pre_par) %>%
    summarize(meantisp=mean(tisp), .groups='keep') %>%
    ungroup() %>%
    spread(key=pre_par, value=meantisp, sep='=')

```


```{r echo=FALSE, include=FALSE}

predictionFunc <- function(fld) {
    
    # Working copy of df, remove generated vars
    df1a <- df1 %>%
        mutate(year=as.numeric(as.character(year))) %>%
        dplyr::select(-case_id, -state_id, -lag_incr) %>%
        dplyr::select(!starts_with('d_')) %>%
        dplyr::select(!starts_with('cop_')) %>%
        dplyr::select(!starts_with('welf_')) %>%
        dplyr::select(!starts_with('unemp_')) %>%
        dplyr::select(!starts_with('pov_')) %>%
        dplyr::select(!starts_with('black_'))
    df1a <- df1a[,1:148]
    
    # Replace missing values
    for (i in 1:nrow(dfvar)) {
        v <- dfvar[i, 'var']
        m <- dfvar[i, 'missing']
        #print(paste0(v, ': missing=', m))
        if (v %in% colnames(df1a)) {
            df1a[[v]] = ifelse(df1a[[v]]==m, NaN, eval(df1a[[v]]))
        }
    }
    
    # Choose only complete cases for the outcome var
    df1a <- df1a[complete.cases(df1a[, c(fld)]),]
    
    # Omit columns with NaN
    df1a <- df1a %>%
        select_if(~ !any(is.na(.)))
    
    # Hypothetical -- no tis laws passed -- use values pre 1994
    df1h <- df1a %>%
        filter(year < 1994 & year > 1972)

    # Predicted values
    df1p <- df1a %>%
        filter(year >= 1994) %>%
        dplyr::select(-sym(fld))
    
    # Model using pre-1994 data
    fm <- as.formula(paste(fld, '~ .'))
    lmod_full <- lm(fm, data=df1h, na.action=na.omit)
    #summary(lmod_full)
    lmod_step1 <- stepAIC(lmod_full, direction='forward', trace=T)
    summary(lmod_step1)
    df1p[fld] <- predict(lmod_step1, newdata=df1p)

    # Recombine data
    df1h <- df1h %>%
        dplyr::select(state_na, year, sym(fld)) %>%
        mutate(val_type='actual_pre-1994')
    dftmp1a <- df1a %>%
        filter(year >= 1994) %>%
        dplyr::select(state_na, year, sym(fld)) %>%
        mutate(val_type='actual_post-1994')
    df1h <- rbind(df1h, dftmp1a)
    df1p <- df1p %>%
        dplyr::select(state_na, year, sym(fld)) %>%
        mutate(val_type='predicted_post-1994')
    df1b <- rbind(df1h, df1p)
    
    # Return
    return(df1b)

}

# Predictions
dfpred_tmp <- predictionFunc('inc_rate')  # incarceration rate
dfpred <- dfpred_tmp
dfpred_tmp <- predictionFunc('cri_l1')  # crime rate
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('admissio')  # prison admissions
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('adm_100k')  # prison admissions per 100k residents
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('ncourtc')  # new court commitments
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('court_10')  # court commitments per 100k residents
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('drug_arr')  # drug arrests per capita
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('pro_drug')  # drug arrests 3-yr avg
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('vio_l1')  # violent crime rate
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)
dfpred_tmp <- predictionFunc('pro_crim')  # property crime rate
dfpred <- dfpred %>% merge(dfpred_tmp, all.x=T)

# Save to file
write.csv(dfpred, 'C:/Users/micha/Box Sync/cuny/608-vis-analytics/project/data/predicted.csv')

```

```{r echo=FALSE, fig.height=3}

# Violent crime
dfpred %>%
    group_by(year, val_type) %>%
    summarize(mean_rate=mean(vio_l1), .groups='keep') %>%
    ggplot(aes(x=year, y=mean_rate, color=val_type)) +
    geom_point(stat='identity') +
    geom_line() +
    theme_minimal() +
    ggtitle("Actual and predicted violent crime rates, pre- and post-1994")

```

But opponents of TIS point out that this is at the expense of growing the prison population; as the data shows, the incarceration rate seems to have climbed under TIS:

```{r echo=FALSE, fig.height=3}

# Incarceration rate
dfpred %>%
    group_by(year, val_type) %>%
    summarize(mean_rate=mean(inc_rate), .groups='keep') %>%
    ggplot(aes(x=year, y=mean_rate, color=val_type)) +
    geom_point(stat='identity') +
    geom_line() +
    theme_minimal() +
    ggtitle("Actual and predicted incarceration rates, pre- and post-1994")

```

So answering the question of whether TIS is working is challenging and may come down to perspective.

Regardless of ideology, TIS is an important topic for not only those it directly touches but for society in general, as we ask what the "right" amount of law enforcement is and the cost of such enforcement in terms of both dollars and the number of people we send to prison.


## References

Stemen, D. _Impact of State Sentencing Policies on Incarceration Rates in the United States, 1975-2002 (ICPSR 4456)._ (2007, September 27). Inter-university Consortium for Political and Social Research. https://www.icpsr.umich.edu/web/NACJD/studies/4456

Ditton, P., Wilson, D. & BJS Statisticians. _Truth in Sentencing in State Prisons._ (1999, January). Bureau of Justice. https://bjs.ojp.gov/content/pub/pdf/tssp.pdf

Shorey, J. _Truth in Sentencing Overview & Laws._ (2022, May 4). Study.com. https://study.com/learn/lesson/truth-in-sentencing-overview-laws.html

_State Good Time and Earned Time Laws (2021, June 11)._ National Conference of State Legislatures. https://www.ncsl.org/civil-and-criminal-justice/state-good-time-and-earned-time-laws

_Know More: Truth-in-Sentencing._ (2020). Restore Justice Foundation. https://www.restorejustice.org/about-us/resources/know-more/know-more-truth-in-sentencing/

_Federal Sentencing Guidelines: Background, Legal Analysis, and Policy Options_ (2009, March 16). EveryCRSReport. https://www.everycrsreport.com/reports/RL32766.html

_Metadata Glossary_ (2023). World Bank, https://databank.worldbank.org/metadataglossary/gender-statistics/series/SI.POV.GINI.
